\textbf{quicR}:\ An R Library for Streamlined Data Handling of Real-Time Quaking Induced Conversion Assays

### 2025-05-22 man/manuscript/softwareX_submission.tex
Added benchmarking for the rtquicR package as well.

        \begin{keyword}
            R package \sep{} RT-QuIC \sep{} Nano-QuIC \sep{} Micro-QuIC \sep{} prion \sep{} diagnostics \sep{} Creutzfeldt-Jakob \sep{} Chronic Wasting Disease \sep{} Parkinson's
        \end{keyword}

@@

    \section{Motivation and significance}
        Real-time quaking induced conversion (RT-QuIC) is type of seed amplification assay (SAA) that has garnered significant attention for its ultra-sensitive detection of misfolded protein aggregates \cite{Wilham2010, Atarashi2011}. The assay works by converting a recombinant protein substrate into an amyloid aggregate in the presence of a misfolded seed \cite{Wilham2010, Orru2012, Orru2017, Orru2015, Bongianni2019, Dassanayake2016, Hwang2018, Groveman2018, Metrick2020}. The assay's sensitivity and specificity make RT-QuIC a promising tool for diagnosing diseases such as prion disorders and other protein misfolding pathologies \cite{Fiorini2020, Franceschini2017, Picasso-Risso2022, Holz2021}. However, the relatively recent development and novelty of the assay have left a gap in widely accepted academic and industry standards for data analysis and interpretation \cite{Rowden2023}.

@@

        \subsection{Comparable Methods}
            To contextualize the functionality and performance of \textbf{quicR}, we compared it with existing RT-QuIC analysis approaches from the published R package, \textbf{QuICSeedR} \cite{Li2025}. While both \textbf{quicR} and \textbf{QuICSeedR} offer overlapping core functionalities, their design philosophies, input handling, and computational trade-offs differ substantially.

            Benchmarking was performed using equivalent functions between the two packages, and results are summarized in Table~\ref{tbl:benchmark}. In general, \textbf{QuICSeedR} out-performs \textbf{quicR} across the board. However, \textbf{quicR} addresses several practical limitations that the authors believe justifies the slower times. Notably, \textbf{QuICSeedR} requires an additional, manually formatted Excel file to input sample metadata, whereas \textbf{quicR} directly parses raw and meta data embedded within a single MARS export file. This design reduces user overhead and minimizes potential for transcription errors.

            \textbf{quicR} also provides more granular control over metric calculations, allowing users to compute MPR, MS, TtT, and RAF independently rather than as a fixed batch. While \textbf{QuICSeedR} includes built-in support for bulk analysis and diagnostic outputs, these features can be approximated in \textbf{quicR} via standard R idioms (e.g., \textbf{apply} family of functions), and are planned for future development.

            Importantly, \textbf{quicR} is actively maintained, in contrast to \textbf{QuICSeedR}, which, to the authors’ knowledge, is no longer under active development. Overall, \textbf{quicR} favors flexibility, reproducibility, and usability with raw instrument output at the expense of execution speed—trade-offs that align with the package’s intended use in diagnostic pipelines.

            \begin{table}[ht]
                \centering
                \begin{tblr}{
                    cells     = {font = \fontsize{11pt}{11pt}\selectfont},
                    colspec   = {rcc}, 
                    row{1}    = {font=\bfseries},
                    rowhead   = 1,
                    width     = \textwidth
                }
                    \hline
                    \SetCell[c=1]{c}Feature / Benchmark & \textbf{quicR} & \textbf{QuICSeedR} \\ 
                    \hline
                    Number of input files required & 1 & 2 \\ 
                    Number of threshold methods & 2 & 3 \\
                    Automated metadata entry & Yes & No \\
                    Actively maintained & Yes & No \\
                    Built-in bulk analysis support & No & Yes \\
                    Diagnostic analysis functionality & No & Yes \\
                    Incorporates dilution factors & Yes & No \\
                    Converts time units & No & Yes \\
                    Calculates metrics independently & Yes & No \\
                    \hline
                    Plate view plot execution time (s) & 0.0978 & 0.0616 \\
                    Metric plot execution time (s) & 0.0108 & 0.0052 \\
                    Metric calculations (MPR, MS, etc.) (s) & 0.3912 & 0.0184 \\
                    Metadata parsing time (s) & 0.1376 & 0.0374 \\
                    Raw data parsing time (s) & 0.0722 & 0.0498 \\
                    Bulk analysis (6 files) (s) & 10.9738 & 0.8486 \\
                    \hline
                \end{tblr}
                \caption{Comparison of core features and execution times between \textbf{quicR} and \textbf{QuICSeedR}. While \textbf{QuICSeedR} demonstrates superior execution speed, \textbf{quicR} offers enhanced metadata handling, metric flexibility, and support for unprocessed instrument output files. All benchmarks were performed using the same dataset and machine (Windows 11, 16GB RAM).}
                \label{tbl:benchmark}
            \end{table}

>>>>

        \begin{keyword}
            Chronic Wasting Disease \sep{} Creutzfeldt-Jakob Disease \sep{} Nano-QuIC \sep{} RT-QuIC \sep{} prion diagnostics \sep{} Parkinson's Disease \sep{} seed amplification assay
        \end{keyword}

@@

    \section{Motivation and significance}
        Real-time quaking induced conversion (RT-QuIC) is within the family of seed amplification assays (SAAs), similiar to protein misfolding cyclic amplification (PMCA), and has garnered significant attention for its ultra-sensitive detection of misfolded protein aggregates \cite{Wilham2010, Atarashi2011}. The assay works by converting a recombinant protein substrate into an amyloid aggregate in the presence of a misfolded seed \cite{Wilham2010, Orru2012, Orru2017, Orru2015, Bongianni2019, Dassanayake2016, Hwang2018, Groveman2018, Metrick2020}. The assay's sensitivity and specificity make RT-QuIC a promising tool for diagnosing diseases such as prion disorders and other protein misfolding pathologies \cite{Fiorini2020, Franceschini2017, Picasso-Risso2022, Holz2021}. However, the relatively recent development and novelty of the assay have left a gap in widely accepted academic and industry standards for data analysis and interpretation \cite{Rowden2023}.

@@

        \subsection{Comparable Methods}
            To contextualize the functionality and performance of \textbf{quicR}, we compared it with existing RT-QuIC analysis approaches from the published R packages, \textbf{QuICSeedR} \cite{Li2025} and \textbf{rtquicR} \cite{Slota2023}. While all three offer overlapping core functionalities, their design philosophies, input handling, and computational trade-offs differ substantially.

            Benchmarking was performed using the rbenchmark library \cite{rbenchmark} with 50 replications for each function. Comparisons are approximate as equivalent functions were not necessarily found between the three packages, and results are summarized in Table~\ref{tbl:benchmark}. In general, \textbf{QuICSeedR} out-performs \textbf{quicR} across the board. However, \textbf{quicR} addresses several practical limitations that the authors believe justifies the slower times. Notably, \textbf{QuICSeedR} requires an additional, manually formatted Excel file to input sample metadata, whereas \textbf{quicR} directly parses raw and meta data embedded within a single MARS export file. This design reduces user overhead and minimizes potential for transcription errors.

            \textbf{quicR} also provides more granular control over metric calculations, allowing users to compute MPR, MS, TtT, and RAF independently rather than as a fixed batch. While \textbf{QuICSeedR} includes built-in support for bulk analysis and diagnostic outputs, these features can be approximated in \textbf{quicR} via standard R idioms (e.g., `apply' family of functions), and are planned for future development.

            Overall, \textbf{quicR} favors flexibility, reproducibility, and usability with raw instrument output at the expense of execution speed—trade-offs that align with the package’s intended use in diagnostic pipelines.

            \begin{table}[H]
                \centering
                \begin{tblr}{
                    cells     = {font = \fontsize{11pt}{11pt}\selectfont},
                    colspec   = {rccc}, 
                    row{1}    = {font=\bfseries},
                    rowhead   = 1,
                    width     = \textwidth
                }
                    \hline
                    \SetCell[c=1]{c}Feature / Benchmark & \textbf{quicR} & \textbf{QuICSeedR} & \textbf{rtquicR} \\ 
                    \hline
                    Available on CRAN & Yes & Yes & No \\
                    Number of input files required & 1 & 2 & 2 \\ 
                    Number of threshold methods & 2 & 3 & 4 \\
                    Number of normalization methods & 1 & 1 & 2 \\
                    Automated metadata entry & Yes & No & No \\
                    Automated sample locations & Yes & Yes & No \\
                    Allows empty data & Yes & Yes & No \\
                    Compatible with 384-well plates & Yes & Yes & No \\
                    % Actively maintained & Yes & No \\
                    Built-in bulk analysis support & No & Yes & Yes \\
                    Diagnostic analysis functionality & No & Yes & No \\
                    Incorporates dilution factors & Yes & No & Yes \\
                    Converts time units & No & Yes & Yes \\
                    Calculates metrics independently & Yes & No & Yes \\
                    Calculates metrics in bulk & Yes & Yes & No \\
                    Calcualates MPR & Yes & Yes & Yes \\
                    Calculates MS & Yes & Yes & No \\
                    Calculates TtT & Yes & Yes & Yes \\
                    Calculates RAF & Yes & Yes & No \\
                    Calculates area under the curve & No & No & Yes \\
                    Calculates SD50 & No & No & Yes \\
                    \hline
                    Plate view plot execution time (s) & 0.082 & 0.062 & NA \\
                    Metric plot execution time (s) & 0.009 & 0.005 & 0.007 \\
                    Metric calculations (MPR, MS, etc.) (s) & 0.355 & 0.018 & NA\\
                    MPR calculation (s) & 0.002 & NA & 0.028 \\
                    TtT calculation (s) & 0.001 & NA & 0.131 \\
                    Metadata parsing time (s) & 0.127 & 0.037 & NA \\
                    Raw data parsing time (s) & 0.069 & 0.050 & 0.191 \\
                    Bulk analysis (s/file) & 6.550 & 0.849 & 1.986 \\
                    \hline
                \end{tblr}
                \caption{Comparison of core features and execution times between \textbf{quicR} and \textbf{QuICSeedR}. While \textbf{QuICSeedR} demonstrates superior execution speed, \textbf{quicR} offers enhanced metadata handling, metric flexibility, and support for unprocessed instrument output files. All benchmarks were performed using the same dataset and machine (Windows 11, 16GB RAM).}
                \label{tbl:benchmark}
            \end{table}
